---
title: 'A4: Incarceration'
author: "Leila Peitsch"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(dplyr)
library(plotly)
library(GGally)
library(usmap)
```

### Introduction

In the US, black people have been found to be overrepresented in jails and
prisons despite only making up [12% of local populations](https://www.pewtrusts.org/en/research-and-analysis/issue-briefs/2023/05/racial-disparities-persist-in-many-us-jails). 
A everlasting trend within American prisons is young black men making up a 
majority of all incarcerations. Following the BLM movement resurgence in 2020 as
a reaction to the murder of George Floyd committed by the police - more
awareness has been brought to the unjust treatment of black individuals by
the police and overall justice system. In this report, I will be looking at data
covering the US jail populations of different races, specifically that of black
and white populations, and how race, gender, age, and location can affect the
demographics within the jail.

### Summary Information

```{r summary, echo = FALSE, message = FALSE}
source("summary.R")
```

When observing the demographics with those incarcerating in local jails, it is
important to keep in mind how the individuals' race, age, and location can
affect how these statistics change between state lines. For example, the state
with the highest jail population within urban neighborhoods (defined as one of 
the core counties of a metropolitan area with about a million people) was 
`r high_urban_state_jail_pop`.
However, the state with the highest jail population within rural neighborhoods 
(defined as non-metro areas with less than 10,000 people) was 
`r high_rural_state_jail_pop`.
Based on state demographics, `r high_rural_state_jail_pop` has the highest 
black population in the US, despite the group only making up [12% of the total population](https://blackdemographics.com/population/black-state-population/).
Looking at the jail populations across races in Texas, specifically that of 
white and black populations, the highest reported 

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)


### Trends Over Time Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The first chart that you will create and include will show the trend over time of your variable/topic. Think carefully about what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Show more than one, but fewer than ~10 trends
  - This may mean showing the same measure for different locations or different racial groups. Think carefully about a meaningful comparison of locations (e.g., the top 10 counties in a state, top 10 states, etc.)
- You must have clear x and y axis labels
- The chart needs a clear title
- You need a legend for your different line colors and a clear legend title

When we say "clear" or "human readable" titles and labels, that means that you should not just display the variable name.

Here's an example of how to run an R script inside an RMarkdown file:

```{r pop over time, echo = FALSE, message = FALSE}
source("chart1_example.R")
plot(plot1)
```

### Variable Comparison Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The second chart that you will create and include will show how two different (continuous) variables are related to one another. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- You must have clear x and y axis labels
- The chart needs a clear title
- If you choose to add a color encoding (not required), you need a legend for your different color and a clear legend title

```{r pop scatterplot, echo = FALSE, message = FALSE}
source("chart2_example.R")
print(plot2)
```

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

```{r map, echo = FALSE, message = FALSE}
source("map_example.R")
plot(map)
```